### 2022-07-14 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@1b21aac669c69434820e120640ea1bea88b05e52 🚀

Statistical and mathematical models can help fill many gaps in our
knowledge about species interactions. Two complementary types of models
have been developed in network ecology for this purpose. On one hand,
predictive models can partially alleviate the Eltonian shortfall, which
describes our current lack of knowledge on food webs and other
ecological networks (Hortal et al. 2015). A variety of such models have
recently been developed using machine learning and other statistical
tools, most of which are presented in Strydom et al. (2021). On the
other hand, null models help us identify potential ecological mechanisms
that drive species interactions. They do so by comparing empirical data
with an unbiased distribution of measures generated using a set of rules
that exclude the mechanism of interest (M. A. Fortuna and Bascompte
2006; Delmas et al. 2019). Both types of models are frequently
topological, i.e.~they often predict the adjacency matrix or specific
measures of network structure without taking into account species'
identity. According to Strydom et al. (2021), these topological models
could be used to make better predictions of pairwise species
interactions by constraining the space of feasible networks.

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. This mathematical method, briefly presented
in Box 1, has been used in a wide range of disciplines, from
thermodynamics to chemistry and biology (Martyushev and Seleznev 2006).
It has also been proven useful in ecology, e.g.~in species distribution
models (Phillips, Anderson, and Schapire 2006) and macroecological
models (J. Harte et al. 2008; John Harte and Newman 2014). As discussed
in Box 1, maximizing a measure of entropy ensures that the derived
probability distributions are unique and least biased under the set of
constraints used. These constraints are built using state variables,
i.e.~variables that represent the macrostate of the system. The
challenge is to find the set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make rigorous predictions using a minimal amount of
data, as well as helping us describe the most important factors driving
that system.

@@

The principle of maximum entropy is a mathematical method of finding
probability distributions, strongly rooted in statistical mechanics and
information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
Starting from a set of constraints given by prior knowledge of a system
(i.e., what we call state variables), this method helps us find
least-biased probability distributions subject to the constraints. These
probability distributions are guaranteed to be unique given our prior
knowledge and represent the most we can say about a system without
making more assumptions. For example, if the only thing we know about a
biological community is its average number of individuals per species,
the least-biased inference we could make on its species abundance
distribution is the exponential distribution (Frank and Smith 2011; John
Harte and Newman 2014). However, this does not imply that this
distribution will be the best fit to empirical data. The challenge is to
find the right set of constraints that would best reproduce
distributions found in nature.

@@

Our maximum entropy network models (second approach only) were compared
with two topological null models. The first is the type I null model of
M. A. Fortuna and Bascompte (2006), in which the probability that a
species \(i\) predates on another species \(j\) is given by

@@

We compared different measures of the structure of empirical food webs
with the ones of null, neutral, and MaxEnt networks
tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
Nestedness indicates how much the diet of specialist species is a subset
of the one of generalists, and is strongly associated with network
modularity (Miguel A. Fortuna et al. 2010). We also measured the maximum
trophic level \emph{maxtl}, network diameter \emph{diam} (i.e., the
longest of the shortest paths between all species pairs; Albert and
Barabasi 2002), the average maximum similarity between species pairs
\emph{MxSim}, the proportion of cannibal species \emph{Cannib} (i.e.,
the proportion of self loops), and the proportion of omnivorous species
\emph{Omniv} (i.e., species whose preys are of different trophic
levels). \emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply
defined in Williams and Martinez (2000).

@@

\leavevmode\hypertarget{ref-Fortuna2010NesMod}{}%
Fortuna, Miguel A., Daniel B. Stouffer, Jens M. Olesen, Pedro Jordano,
David Mouillot, Boris R. Krasnov, Robert Poulin, and Jordi Bascompte.
2010. {``Nestedness Versus Modularity in Ecological Networks: Two Sides
of the Same Coin?''} \emph{Journal of Animal Ecology} 79 (4): 811--17.
\url{https://doi.org/10.1111/j.1365-2656.2010.01688.x}.

@@

@@

\leavevmode\hypertarget{ref-Harte2014MaxInf}{}%
Harte, John, and Erica A. Newman. 2014. {``Maximum Information Entropy:
A Foundation for Ecological Theory.''} \emph{Trends in Ecology \&
Evolution} 29 (7): 384--89.
\url{https://doi.org/10.1016/j.tree.2014.04.009}.

>>>>

Statistical and mathematical models can help fill many gaps in our
knowledge about species interactions. Two complementary types of models
have been developed in network ecology for this purpose. On one hand,
predictive models can partially alleviate the Eltonian shortfall, which
describes our current lack of knowledge on food webs and other
ecological networks (Hortal et al. 2015). A variety of such models have
recently been developed using machine learning and other statistical
tools, most of which are presented in Strydom et al. (2021). On the
other hand, null models help us identify potential ecological mechanisms
that drive species interactions. They do so by comparing empirical data
with an unbiased distribution of measures generated using a set of rules
that exclude the mechanism of interest (Fortuna and Bascompte 2006;
Delmas et al. 2019). Both types of models are frequently topological,
i.e.~they often predict the adjacency matrix or specific measures of
network structure without taking into account species' identity.
According to Strydom et al. (2021), these topological models could be
used to make better predictions of pairwise species interactions by
constraining the space of feasible networks.

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. This mathematical method, briefly presented
in Box 1, has been used in a wide range of disciplines, from
thermodynamics to chemistry and biology (Martyushev and Seleznev 2006).
It has also been proven useful in ecology, e.g.~in species distribution
models (Phillips, Anderson, and Schapire 2006) and macroecological
models (Harte et al. 2008; Harte and Newman 2014). As discussed in Box
1, maximizing a measure of entropy ensures that the derived probability
distributions are unique and least biased under the set of constraints
used. These constraints are built using state variables, i.e.~variables
that represent the macrostate of the system. The challenge is to find
the set of state variables that best represent natural systems and to
translate them into appropriate statistical constraints. Having a
validated maximum entropy model for the system at hand allows us to make
rigorous predictions using a minimal amount of data, as well as helping
us describe the most important factors driving that system.

@@

The principle of maximum entropy is a mathematical method of finding
probability distributions, strongly rooted in statistical mechanics and
information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
Starting from a set of constraints given by prior knowledge of a system
(i.e., what we call state variables), this method helps us find
least-biased probability distributions subject to the constraints. These
probability distributions are guaranteed to be unique given our prior
knowledge and represent the most we can say about a system without
making more assumptions. For example, if the only thing we know about a
biological community is its average number of individuals per species,
the least-biased inference we could make on its species abundance
distribution is the exponential distribution (Frank and Smith 2011;
Harte and Newman 2014). However, this does not imply that this
distribution will be the best fit to empirical data. The challenge is to
find the right set of constraints that would best reproduce
distributions found in nature.

@@

Our maximum entropy network models (second approach only) were compared
with two topological null models. The first is the type I null model of
Fortuna and Bascompte (2006), in which the probability that a species
\(i\) predates on another species \(j\) is given by

@@

We compared different measures of the structure of empirical food webs
with the ones of null, neutral, and MaxEnt networks
tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
Nestedness indicates how much the diet of specialist species is a subset
of the one of generalists, and is strongly associated with network
modularity (Fortuna et al. 2010). We also measured the maximum trophic
level \emph{maxtl}, network diameter \emph{diam} (i.e., the longest of
the shortest paths between all species pairs; Albert and Barabasi 2002),
the average maximum similarity between species pairs \emph{MxSim}, the
proportion of cannibal species \emph{Cannib} (i.e., the proportion of
self loops), and the proportion of omnivorous species \emph{Omniv}
(i.e., species whose preys are of different trophic levels).
\emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply defined in
Williams and Martinez (2000).

@@

\leavevmode\hypertarget{ref-Fortuna2010NesMod}{}%
Fortuna, M. A., Daniel B. Stouffer, Jens M. Olesen, Pedro Jordano, David
Mouillot, Boris R. Krasnov, Robert Poulin, and Jordi Bascompte. 2010.
{``Nestedness Versus Modularity in Ecological Networks: Two Sides of the
Same Coin?''} \emph{Journal of Animal Ecology} 79 (4): 811--17.
\url{https://doi.org/10.1111/j.1365-2656.2010.01688.x}.

@@

\leavevmode\hypertarget{ref-Harte2014MaxInf}{}%
Harte, J., and Erica A. Newman. 2014. {``Maximum Information Entropy: A
Foundation for Ecological Theory.''} \emph{Trends in Ecology \&
Evolution} 29 (7): 384--89.
\url{https://doi.org/10.1016/j.tree.2014.04.009}.

@@
